Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
discussion p2

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of
GD-1, for the first time we find evidence of stream members off of the main
track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

The absence of its progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 (even at its current length of $\sim90^\circ$) a top modeling priority to map the inner 30\,kpc of the Galaxy.
Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.

- spur \& blob as related to gaps? -- classify possible explanations in this way
- if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v

* no longer simple stream, discuss possible causes

- progenitor
-- rotation -- could explain spurs, but not gaps?

- smooth component of the potential?
-- show orbit
-- chaos?
-- \todo{APW} bar -- hard to make small feature, size should be ~angular size of the bar at stream (influence of bar should be a radian of the stream) + relative velocity (retrograde)

- encounters?
-- work out mass / time combinations in impulse approximation
-- comment on GMC vs subhalo

- future:
-- radial velocities, measure 3d velocity structure (unique predictions)
- delta energy,
- cleaner selection (any stars in the gap)

- (should go at least to symmetric in positive), not captured with naive selection, as there are gradients
- prediction: very negative rv, should be easy to map the full extent

% -- tests for other outstanding issues
-- model facets of stream morphology to ascertain if consistent with encounter signatures
-- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
of stream members off of the main track.
These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

The absence of its progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 a top modeling priority to map the inner 30\,kpc of the Galaxy, even at the length of $\sim90^\circ$ inferred in this work.
Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.

GD-1 seen by \gaia\ is no longer a simple, one-dimensional structure.
Not only are the density variations, tentatively deduced in previous studies, confirmed as highly significant, but for the first time we have mapped tidal debris from a globular cluster beyond a thin stellar track: parts of GD-1 are at a distance of up to $\sim1^\circ$ from the main track.
Separation of a part of a stellar stream from the main track could be caused by the different exit velocity from the cluster, but explaining the amount of separation and velocity offset in GD-1 would require the progenitor to rotate at XX km\,s$^{-1}$, much faster than measured for the known clusters (cit).
Features off from the cold part of the stream are likely too small to be caused by chaotic dispersal from either triaxial matter distribution (cit apw) or the time-dependent influence of the rotating bar (cit pearson).
However, encounters with a massive perturber are expected to open gaps in streams (cits), so we might be seeing the expelled material in the process of phase mixing away from the stream track.
The encounter origin would be especially exciting -- since GD-1 resides in the halo, the only objects massive enough likely to cross its path are the long-searched-after dark-matter subhalos.

Finding the origin of gaps and off-track features in GD-1 will require progress on two fronts: theoretically, we need to understand the co-evolution of the complex stream system under different origin scenarios, while observing radial velocities of candidate GD-1 members would facilitate data interpretation by remove the remaining contamination and enabling studies in the space of conserved quantities.
-- sentence on getting more from off-track than gaps alone, but somehow more general and not tied specifically to the encounter scenario / or some other nice parting thought

% -- radial velocities, measure 3d velocity structure (unique predictions)
% - if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v -> delta energy
% - cleaner selection (any stars in the gap?)
% - prediction: very negative rv, should be easy to map the full extent

% -- model facets of stream morphology to ascertain if consistent with encounter signatures
% -- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history
